Internal Gravity Wave Breaking in White Dwarf Binaries

### 2019-11-13 writeups/Su_IGW_break.tex
wtf ok actual last
updated a ton of plots, first real draft

    \documentclass[
        fleqn,
        usenatbib,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        ae, aecompl,
        graphicx,
        booktabs,
    }
    \usepackage[T1]{fontenc}
    \usepackage[
        labelfont=bf, % caption names are labeled in bold
        font=scriptsize % smaller font for captions
    ]{caption}
    \usepackage[caption=false]{subfig} % subfigures

@@

\begin{abstract}
    In sufficiently compact white dwarf binaries, dynamical tides raise a train
    of internal gravity waves that propagate towards the surface. We perform 2D
    numerical simulations of these waves undergoing nonlinear wave breaking in
    an incompressible, isothermal atmosphere. After an initial transient phase,
    we find that these waves induce a sharp transition between a non-rotating
    core and synchronously rotating envelope. We find evidence that the width of
    this transition layer is bound from below by the Kelvin-Helmholtz
    Instability. We provide analytical formulae for absorption and reflection of
    incident waves off the critical layer up to prefactors of order unity. These
    prefactors converge to constant values when artificial dissipation is
    decreased. We provide dimensionless criteria necessary to resolving momentum
    transfer within the critical layer. Finally, we speculate on the application
    of our model to tidal synchronization and heating in astrophysical systems.
\end{abstract}

@@

[Below copied from proposal, will rewrite]

@@

% Usual WD binaries introduction, from proposals

% IGW tidal dissipation has been thoroughly studied in many astrophysical
% binary systems (Barker Ogilvie, Goldreich Nicholson, Zahn?). Its effect in WDs
% (Fuller Lai) resembles early solar type stars but has not been studied
% numerically. While the linear IGW solution is an exact solution of the
% nonlinear system (Drazin) but is globally unstable (Drazin), the IVP must be
% studied numerically to determine how the linear solution breaks down and
% deposits energy/angular momentum.

% From numerical studies, IGWs are known to be anti-diffusive (Lecoanet mean
% flow papers, Lindzen) and sharpen mean flows (m = 0). Mean flows produce
% reflection, known both analytically (Booker & Bretherton) and numerically
% (Winters d'Asaro). IGW self-interaction via its wave-induced mean flow is
% known to play a crucial role in IGW wavebreaking (Sutherland x2).

% We study IGW breaking in the plane-parallel approximation in 2D. Known to
% describe both wave steepening (klostermeyer 1991) and captures mean flow
% absorption (Booker & Bretherton). While the details of breaking are presently
% understood to be 3D (Winters & d'Asaro), we argue in (section) that our
% results should be qualitatively similar to 3D (both 2D and 3D exhibit KHI, and
% since we are able to resolve KHI, the spinup in 3D should be similar?).

In \autoref{s:equations}, we will describe the system of equations we will
use to analyze IGW breaking. In \autoref{s:theory}, we discuss relevant
analytical results. In \autoref{s:sim} we present the results of numerical
simulations. Finally, in \autoref{s:discussion} we discuss the results of the
preceeding section.

@@

We consider a incompressible, isothermal fluid, representative of degenerate
matter in WDs. We use barotropic equation of state $P(\rho, T) = P(\rho)$ as a
first approximation. As we are interested in dynamics far from the center of the
WD, we approximate the gravitational field as uniform. We model the background
density stratification as $\overline{\rho} = \overline{\rho}_0 e^{-z/H}$ for
some reference density $\overline{\rho}_0$ (we generally notate background
quantities with overbars and perturbation quantities with primes). Finally, we
consider a 2D fluid for computational feasibility: while it is well-known wave
breaking is a 3D process \citep{klostermeyer,winters1994}, the dynamical effect
of the breaking process is likely to be similar in 2D \citep{barker_ogilvie}.

@@

In the small perturbation limit, where flow velocities are small compared to the
characteristic space and time scales $\pd{}{t} \gg \bm{u}' \cdot
\bm{\nabla}$, we may linearize \autoref{se:nl_upsilon}. The solution in this
linear regime is given by \citep{drazin,sutherland0}:
\begin{equation}
    u'_z\p{x, z, t} = Ae^{z/2H}\cos\p{k_{x}x + k_{z}z - \omega t},
        \label{eq:k0z_sign}
\end{equation}
where $A$ is the amplitude and $\omega$ satisfies dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}.
        \label{eq:disp_rel}
\end{equation}
Our equations are valid in the limit of large sound speed $c_s \to \infty$, in
which the \emph{Brunt-V\"ais\"al\"a frequency}
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
is constant. Other dynamical quantities are simply related to $u'_z$ \citep[see
e.g.][]{sutherland0}.

@@

    \item The averaged horizontal momentum flux $F$ is
        \begin{align}
            F &\equiv \ev{\rho u_{x}' u_{z}'}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{x}'u_{z}'\;\mathrm{d}x,
                    \label{eq:F_def}\\
                &\approx -\frac{A^2}{2}\overline{\rho}_0\frac{k_{z}}{k_{x}},
                    \label{eq:S_lin}
        \end{align}
        Thus, indeed $F > 0$ for an upward propagating IGW $c_{g, z} > 0$.
\end{itemize}

@@

To model continuous excitation of IGWs deep in the WD interior propagating
towards the surface, we use a volumetric forcing term to excite IGW near the
bottom of the simulation domain. Our forcing excites both IGWs
propagating upwards, imitating a wave tidally excited deeper in the WD, and
downwards, which are damped away by the damping layers described in
\autoref{ss:damping}.

As not to interfere with the incompressibility constraint, we force the system
on the density equation. We implement forcing with strength $C$ localized around
$z_0$ with small width $\sigma$ by replacing \autoref{eq:nl_up_density} with
\begin{equation}
    \md{\Upsilon}{t} + u_{z}'\pd{\ln \overline{\rho}}{z}
        = Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p{k_{x}x - \omega t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $z$ power spectrum, but only the
$k_{z}$ satisfying dispersion relation \autoref{eq:disp_rel} for the given
$k_{x}$, $\omega(k_{x}, k_{z})$ will propagate.

In the linearized system, the effect of this forcing can be solved exactly. If
we approximate $k_zH \gg 1, \sigma \ll H$, the solution can be approximated as
two plane waves propagating away from the forcing zone
\begin{align}
    u_{z}'&(x, z, t) \approx{} \frac{C}{2k_z}\frac{gk_x^2}{\omega^2}
        e^{-\frac{k_z^2\sigma^2}{2}}
        \sqrt{2\pi \sigma^2} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}\sin\p{k_{x}x + k_{z}(z - z_0) - \omega t
            + \frac{k_z\sigma^2}{2H}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}\sin\p{k_{x}x - k_{z}(z - z_0) - \omega t
            + \frac{k_z\sigma^2}{2H}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ region models an upward propagating IGW wavetrain excited deep
in the atmosphere. A derivation of this result is provided in
\autoref{s:force_solved}.

@@

Above, we neglected advective terms $\pd{}{t} \gg \bm{u}' \cdot \bm{\nabla}$.
However, since $\abs{\bm{u}'} \propto e^{z/2H}$ in an infinite domain all IGWs
eventually grow to nonlinear amplitudes and break. Nonlinear wave breaking is
expected to be important in WDs \citep{fullerI,fullerII}. We can
estimate the height of wave breaking using $\frac{\abs{\bm{u}'}
\abs{\bm{k}}}{\omega} \sim 1$. This can be rewritten using the Lagrangian
displacement $-i\omega \bm{\xi}' = \bm{u}'$:
\begin{equation}
    \xi_z k_z \gtrsim 1.\label{eq:nl}
\end{equation}

@@

A horizontal shear flow $\overline{U}(z, t)\hat{x}$ enters the fluid equations
via the Lagrangian derivative, which can be decomposed as
\begin{equation}
    \md{}{t} = \pd{}{t} + \overline{U} \pd{}{x} + \p{\bm{u}' \cdot \vec{\nabla}},
\end{equation}
where $\bm{u}'$ is the velocity field \emph{without} the shear flow, for
clarity. Thus, $\overline{U}$ has the effect of Doppler shifting the time
derivative into the frame comoving with the mean flow. If $\overline{U}$ is
roughly constant, then the behavior of a small plane wave perturbation satisfies
modified dispersion relation
\begin{equation}
    \p{\omega - \overline{U}k_x}^2 =
        \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}.
        \label{eq:disp_rel_U}
\end{equation}
This is just \autoref{eq:disp_rel} with $\omega \to \omega - \overline{U}k_x$.

It is apparent that if $\overline{U} = \overline{U}_c$, where
\begin{equation}
    \overline{U}_c \equiv \frac{\omega}{k_x},
\end{equation}
then the dispersion relation is singular and the linear solution breaks down.
Physically, this corresponds to the Doppler-shifted frequency of the IGW being
zero. Anywhere $\overline{U} = \overline{U}_c$ is called a \emph{critical
layer}.

@@

Consider the horizontal momentum transfer at the critical layer. Any incident
horizontal momentum flux absorbed by the fluid must manifest as additional
horizontal momentum of the shear flow. Since $\overline{U}$ cannot exceed
$\overline{U}_c$, the critical layer will instead propagate downward in response
to the incident momentum flux. The horizontal momentum of the shear flow
satisfies
\begin{equation}
    \pd{}{t}\int\limits \overline{\rho}(z) \overline{U}(z, t)\;\mathrm{d}z
        - \Delta F = 0,
\end{equation}
where $\Delta F > 0$ is the absorbed horizontal momentum. Assuming
$\overline{U}(z > z_c) = \overline{U}_c, \overline{U}(z < z_c) = 0$, this
condition becomes
\begin{equation}
    \overline{\rho}(z_c) \overline{U}_c\rd{z_c}{t} = \Delta F.\label{eq:zc_anal}
\end{equation}
If $\Delta F$ is constant in time, $z_c(t)$ has analytical solution
\begin{equation}
    z_c(t) = -H\ln \s{e^{-\frac{z_c(t = 0)}{H}} +
        \frac{t\Delta F}{\overline{U}_c H\overline{\rho}_0}},\label{eq:zc_sol}
\end{equation}
where $z_c(t = 0)$ is the initial critical layer height.

\section{Numerical Simulation Setup}\label{ss:numerics}

@@

For numerical convenience, we simulate equations \autoref{eq:nl_incomp},
\autoref{eq:nl_upsilon_u}, and \autoref{eq:vol_drive} in a Cartesian box with
size $L_x, L_z$. The $x$ direction represents the azimuthal direction in a star,
and the $z$ direction represents the radial direction. We thus choose periodic
boundary conditions in the $x$ direction. We also use periodic boundary
conditions in the $z$ direction, but we damp perturbations to zero near the
top/bottom of the domain to capture their distinct behavior (described in
\autoref{ss:damping}). We expand all variables as a Fourier series with $N_x$
and $N_z$ modes, and use the $3/2$ dealiasing rule to avoid aliasing errors in
the nonlinear terms \citep{boyd}.

@@

    \item $\nu$: Nonlinear effects transfer wave energy from the injection
        wavenumber $\bm{k}$ to larger wavenumbers. Our spectral method does not
        have any numerical viscosity, so diffusivity must be introduced into the
        equations to regularize the systems. We add viscosity and diffusivity to
        the system in a way that conserves horizontal momentum (see
        Appendix~\ref{se:strat_impl} for details). When the forcing is weak, we
        may set $\nu=0$ because there is negligible energy transfer between
        modes.
\end{itemize}

@@

We aim to damp waves that reach the edge of the simulation domain without
inducing nonphysical reflection. To do so, we replace material derivatives
in \autoref{se:nl_upsilon} with:
\begin{align}
    \md{}{t} &\to \md{}{t} + \Gamma(z)\Upsilon,\\
    \Gamma(z) &= \frac{1}{2\tau}\s{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}},\label{eq:Gamma}
\end{align}
where $z_B = 1.5H, z_T = 10.5H$ are the boundaries of the damping zone.
This strongly damps perturbations below $z_B$ and above $z_T$ with damping time
$\tau$, negligibly affects dynamics between $z_B, z_T$ and has transition width
governed by $\Delta z$. We use $\Delta z = 0.25H$ and $\tau = 1/15N$. This
prescription is similar to \citep{lecoanet_damp} and has the advantage of being
smooth, important for spectral methods. Further details of our implementation of
the fluid equations in Dedalus are described in Appendix~\ref{se:strat_impl}.

@@

We describe the results of a simulation with $C = \scinot{1.64}{-7}$ such that
$\xi_z k_z \ll 1$ everywhere. In particular, $\xi_z k_z \approx \scinot{5}{-5}$
just above the forcing zone.

We expect the waves to follow the solution given by \autoref{eq:uz_lin}. Along
with incompressibility (\autoref{eq:nl_incomp}), we obtain a complete analytical
solution for flow velocities $\bm{u}_{al}(x, z, t)$. The amplitude of the
observed IGW in the simulation field $\bm{u}$ relative to analytical solution
$\bm{u}_{al}$ over some region $z \in [z_b, z_t]$ can be estimated with
estimator $\hat{A}_i(t)$ (subscript $i$ denotes incident wave)
\begin{equation}
    \hat{A}_i(t) = \frac{\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\p{\bm{u} \cdot \bm{u}_{al}}\;\mathrm{d}x\mathrm{d}z}
        {\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\abs{\bm{u}_{al}}^2\;\mathrm{d}x\mathrm{d}z}
        \label{eq:ahat_def}
\end{equation}
If $\bm{u} = \bm{u}_{al}$, then $\hat{A}_i(t) = 1$. The energy norm is a
traditional normalization choice such that the overlap between $\bm{u},
\bm{u}_{al} \propto e^{z/2H}$ is evenly weighted throughout the integration
region.

For a simulation satisfying $\xi_z k_z \ll 1$ everywhere, we expect
$\hat{A}_i(t) = 1$ when integrated between the forcing and damping zones, i.e.\
$z_b \gtrsim z_0, z_t \lesssim z_T$ ($z_0, z_T$ are defined in
\autoref{eq:vol_drive} and \autoref{eq:Gamma} respectively). For consistency
with the nonlinear case later, we choose $z_b = z_0 + 3\sigma, z_t = z_b + H$
(using $z_t = z_T - \Delta z$ just below the damping layer instead does not
change the results). The resulting measurement of $\hat{A}_i(t)$ is shown in
\autoref{fig:lin_amps}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{plots/lin_amps.png}
    \caption{Amplitude of excited wave over time in weak forcing simulation,
    computed using \autoref{eq:ahat_def}. $\hat{A}_i(t) = 1$ corresponds to
    perfect agreement with the analytical estimate. After an initial transient
    phase, we observe great agreement with \autoref{eq:uz_lin}, and in
    particular $\hat{A}_i(t)$ asymptotes to $\approx 1$, implying continuous
    excitation of identical IGW with the expected amplitude. The small
    suppression may be due to timestepping errors, as a relatively large fixed
    step size $\Delta t = 0.1/N$ was used for this
    simulation.}\label{fig:lin_amps}
\end{figure}

@@

% times on fluxes.png are 054, 086, 153

To perform simulations of wave breaking phenomena, we use the same values as
\autoref{ss:numerics} except for $C$ and $\nu$. In particular, we choose $C$
such that $\at{\xi_z k_z}_{z_0} = 0.1$ in the forcing zone. The linear solution
predicts $\xi_z k_z \sim 4.25$ at the upper damping zone $z_T$. The dissipation
parameter $\nu$ was varied across the various simulations. To quantify $\nu$, we
define dimensionless Reynolds number
\begin{equation}
    \mathrm{Re}  \equiv \frac{\omega}{\nu k_{z}^2}. \label{eq:re_def}
\end{equation}
A table of our simulations can be found in \autoref{tab:params}.
\begin{table}
    \centering
    \begin{tabular}{l c c c}
        Resolution & $\mathrm{Re}$\\\bottomrule
        $1024 \times 4096$ & $2048$\\
        $768 \times 3072$ & $1024$\\
        $512 \times 2048$ & $512$\\
        $256 \times 1024$ & $341$\\
        $256 \times 1024$ & $205$\\
        $256 \times 1024$ & $146$\\
    \end{tabular}
    \caption{Table of simulation resolutions.}\label{tab:params}
\end{table}

@@

A full video of our higher-resolution run at $N_x = 768$, $N_z = 3072$,
$\mathrm{Re} = 1024$ is available
online\footnote{http://www.princeton.edu/~lecoanet/data/breaking\_wave.mov}. We
take this to be our fiducial simulation for the remainder of this paper, though
other simulations show qualitatively similar behavior.

>>>>

    \documentclass[
        fleqn,
        usenatbib,
        referee,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        ae, aecompl,
        graphicx,
        booktabs,
    }
    \usepackage[T1]{fontenc}
    \usepackage[
        labelfont=bf, % caption names are labeled in bold
        font=scriptsize % smaller font for captions
    ]{caption}
    \usepackage[caption=false]{subfig} % subfigures

@@

\begin{abstract}
    In sufficiently compact white dwarf binaries, dynamical tides raise a train
    of internal gravity waves that propagate towards the surface and dissipate
    via nonlinear wave breaking. We perform 2D hydrodynamical simulations of
    this wave breaking in an incompressible, isothermal atmosphere. After an
    initial transient phase, we find that these waves generate a sharp
    transition, a critical layer, between the non-rotating core and
    synchronously rotating envelope. We find evidence that indefinite steepening
    of this critical layer is suppressed by the Kelvin-Helmholtz instability. We
    study the absorption and reflection of incident waves off the critical layer
    and provide analytical formulae describing its evolution. We infer
    necessary conditions to resolving momentum transfer within the critical
    layer. Finally, we speculate on the application of our model to tidal
    synchronization and heating in astrophysical systems.
\end{abstract}

@@

@@

In this paper, we perform numerical simulations of tidally-excited IGWs
undergoing wave breaking in an idealized WD fluid. The primary aim of the paper
is to thoroughly characterize the angular momentum transfer within a WD by
purely hydrodynamical effects. As a first approximation, we analyze the problem
in a 2D Cartesian geometry, and consider IGW propagating into a fluid initially
at rest. We find that, after an initial transient phase, the fluid develops two
distinct zones: (i) a lower zone that has no horizontal mean flow, and (ii) an
upper zone with significant horizontal mean flow.

These two zones are separated by a \emph{critical layer}. The majority of our
paper is dedicated to characterizing the behavior of the critical layer when
interacting with a continuous train of IGW excited from the bottom of the fluid.
IGW are generally \emph{anti}-diffusive, in that they steepen shear flows
\citep{lindzen_qbo,lecoanet_meanflow}, and act to narrow the critical layer. We
find this steepening is suppressed by the Kelvin-Helmholtz instability and
turbulence within the very narrow critical layer. By careful accounting of the
momentum flux budget about the critical layer, we are able to model the
propagation of the critical layer, reflection of the incident IGW, and other
momentum redistribution behaviors. Our results are numerically converged
for sufficiently large resolutions.

In \autoref{s:equations}, we will describe the system of equations we will use
to analyze IGW breaking. In \autoref{s:theory}, we review existing understanding
of wave breaking and present new analytical results. In \autoref{s:numerics}, we
describe our numerical setup, which in \autoref{s:weak_sim} we validate in the
weak forcing limit against linear theory. In \autoref{s:sim}, we present the
results of simulation of IGW breaking and our characterization of the critical
layer. Finally, we summarize and conclude in \autoref{s:discussion}.

@@

We consider a incompressible, isothermal fluid, representative of degenerate
matter in WDs. We use barotropic equation of state $P(\rho, T) = P(\rho)$ as a
first approximation. As we are interested in dynamics far from the center of the
WD, we approximate the gravitational field as uniform. We model the background
density stratification as $\overline{\rho} = \overline{\rho}_0 e^{-z/H}$ for
some reference density $\overline{\rho}_0$ (we generally notate background
quantities with overbars and perturbation quantities with primes). We
consider a 2D fluid with for computational feasibility. While
it is well-known wave breaking is a 3D process \citep{klostermeyer,winters1994},
the dynamical effect of the breaking process is likely to be similar in 2D
\citep{barker_ogilvie}. Finally, we simplify by considering simple
plane-parallel IGW\@. Since wave breaking generally occurs near the surface of
WDs \citep[$\gtrsim 0.9R_{WD}$][]{fullerI}, the curvature of the wave front is
negligible.

@@

In the small perturbation limit, where flow velocities are small compared to the
characteristic space and time scales $\pd{}{t} \gg \bm{u} \cdot
\bm{\nabla}$, we may linearize \autoref{se:nl_upsilon}. The solution in this
linear regime is given by \citep{drazin,sutherland0}:
\begin{equation}
    u_z\p{x, z, t} = Ae^{z/2H}\cos\p{k_{x}x + k_{z}z - \omega t},
        \label{eq:k0z_sign}
\end{equation}
where $A$ is the amplitude and $\omega$ satisfies dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}.
        \label{eq:disp_rel}
\end{equation}
Our equations are valid in the limit of large sound speed $c_s \to \infty$, in
which the \emph{Brunt-V\"ais\"al\"a frequency}
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
is constant. Other dynamical quantities are simply related to $u'_z$ \citep[see
e.g.][]{sutherland0}.

@@

    \item The averaged horizontal momentum flux $F$ is
        \begin{align}
            F &\equiv \ev{\rho u_{x} u_{z}}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{x}u_{z}\;\mathrm{d}x,
                    \label{eq:F_def}\\
                &\approx -\frac{A^2}{2}\overline{\rho}_0\frac{k_{z}}{k_{x}},
                    \label{eq:S_lin}
        \end{align}
        Thus, indeed $F > 0$ for an upward propagating IGW $c_{g, z} > 0$.
\end{itemize}

@@

To model continuous excitation of IGWs deep in the WD interior propagating
towards the surface, we use a volumetric forcing term to excite IGW near the
bottom of the simulation domain. Our forcing excites both IGWs
propagating upwards, imitating a wave tidally excited deeper in the WD, and
downwards, which are not physically meaningful, but are damped away by the
damping layers described in \autoref{ss:damping}.

As not to interfere with the incompressibility constraint, we force the system
on the density equation. We implement forcing with strength $C$ localized around
$z_0$ with small width $\sigma$ by replacing \autoref{eq:nl_up_density} with
\begin{equation}
    \md{\Upsilon}{t} + u_{z}\pd{\ln \overline{\rho}}{z}
        = Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p{k_{x}x - \omega t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $z$ power spectrum, but only the
$k_{z}$ satisfying dispersion relation \autoref{eq:disp_rel} for the given
$k_{x}$, $\omega(k_{x}, k_{z})$ will propagate.

In the linearized system, the effect of this forcing can be solved exactly
(\autoref{s:force_solved}). If we approximate $k_zH \gg 1, \sigma \ll H$, the
solution can be approximated as two plane waves propagating away from the
forcing zone
\begin{align}
    u_{z}&(x, z, t) \approx{} \frac{C}{2k_z}\frac{gk_x^2}{\omega^2}
        e^{-\frac{k_z^2\sigma^2}{2}}
        \sqrt{2\pi \sigma^2} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}\sin\p{k_{x}x + k_{z}(z - z_0) - \omega t
            + \frac{k_z\sigma^2}{2H}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}\sin\p{k_{x}x - k_{z}(z - z_0) - \omega t
            + \frac{k_z\sigma^2}{2H}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ region models an upward propagating IGW wavetrain excited deep
in the atmosphere.

@@

Above, we neglected advective terms $\pd{}{t} \gg \bm{u} \cdot \bm{\nabla}$.
However, since $\abs{\bm{u}} \propto e^{z/2H}$ in an infinite domain all IGWs
eventually grow to nonlinear amplitudes and break. Nonlinear wave breaking is
expected to be important in WDs \citep{fullerI,fullerII}. We can
estimate the height of wave breaking using $\frac{\abs{\bm{u}}
\abs{\bm{k}}}{\omega} \sim 1$. This can be rewritten using the Lagrangian
displacement $-i\omega \bm{\xi} = \bm{u}$:
\begin{equation}
    \xi_z k_z \gtrsim 1.\label{eq:nl}
\end{equation}

@@

A horizontal shear flow $\overline{U}(z, t)\hat{x}$ enters the fluid equations
via the Lagrangian derivative, which can be decomposed as
\begin{equation}
    \md{}{t} = \pd{}{t} + \overline{U} \pd{}{x} + \p{\bm{u}' \cdot \vec{\nabla}},
\end{equation}
where $\bm{u}'$ is the velocity field \emph{without} the shear flow, for
clarity. Thus, $\overline{U}$ has the effect of Doppler shifting the time
derivative into the frame comoving with the mean flow. If $\overline{U}$ is
roughly constant, then the behavior of a small plane wave perturbation satisfies
modified dispersion relation
\begin{equation}
    \p{\omega - \overline{U}k_x}^2 =
        \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}.
        \label{eq:disp_rel_U}
\end{equation}
This is just \autoref{eq:disp_rel} with $\omega \to \omega - \overline{U}k_x$.
It is apparent that if $\overline{U} = \overline{U}_c$, where
\begin{equation}
    \overline{U}_c \equiv \frac{\omega}{k_x},
\end{equation}
then the dispersion relation is singular and the linear solution breaks down.
Physically, this corresponds to the Doppler-shifted frequency of the IGW being
zero. Anywhere $\overline{U} = \overline{U}_c$ is called a \emph{critical
layer}.

@@

Consider the horizontal momentum transfer at the critical layer. Any incident
horizontal momentum flux absorbed by the fluid must manifest as additional
horizontal momentum of the shear flow. Since $\overline{U}$ cannot exceed
$\overline{U}_c$, the critical layer will instead propagate downward in response
to the incident momentum flux. The horizontal momentum of the shear flow
satisfies
\begin{equation}
    \pd{}{t}\int\limits \overline{\rho}(z) \overline{U}(z, t)\;\mathrm{d}z
        - F_a(t) = 0,
\end{equation}
where $F_a(t) > 0$ is the absorbed horizontal momentum. Assuming $\overline{U}(z
> z_c) = \overline{U}_c, \overline{U}(z < z_c) = 0$, this condition becomes
\begin{equation}
    \overline{\rho}(z_c) \overline{U}_c\rd{z_c}{t} = F_a(t).\label{eq:zc_anal}
\end{equation}
If $F_a$ is constant in time, $z_c(t)$ has analytical solution
\begin{equation}
    z_c(t) = -H\ln \s{e^{-\frac{z_c(t = 0)}{H}} +
        \frac{tF_a}{\overline{U}_c H\overline{\rho}_0}},\label{eq:zc_sol}
\end{equation}
where $z_c(t = 0)$ is the initial critical layer height.

\section{Numerical Simulation Setup}\label{s:numerics}

@@

We simulate equations \autoref{eq:nl_incomp}, \autoref{eq:nl_upsilon_u}, and
\autoref{eq:vol_drive} in a Cartesian box with size $L_x, L_z$. The $x$
direction represents the azimuthal direction in a star, and the $z$ direction
represents the radial direction. We thus choose periodic boundary conditions in
the $x$ direction. We also use periodic boundary conditions in the $z$
direction, but we damp perturbations to zero near the top/bottom of the domain
to capture their distinct behavior (described in \autoref{ss:damping}). We
expand all variables as Fourier series with $N_x$ and $N_z$ modes, and use the
$3/2$ dealiasing rule to avoid aliasing errors in the nonlinear terms
\citep{boyd}.

@@

    \item $\nu$: Nonlinear effects transfer wave energy from the injection
        wavenumber $\bm{k}$ to larger wavenumbers. Our spectral method does not
        have any numerical viscosity, so diffusivity must be introduced into the
        equations to regularize the systems at large wavenumbers. We add
        viscosity and diffusivity to the system in a way that conserves
        horizontal momentum (see Appendix~\ref{se:strat_impl} for details).

        To quantify $\nu$, we define the dimensionless Reynolds number
        \begin{equation}
            \mathrm{Re}  \equiv \frac{\omega}{\nu k_{z}^2}. \label{eq:re_def}
        \end{equation}
        When the forcing is weak, we may set $\nu \approx 0$, or $\mathrm{Re}
        \gg 1$ because there is negligible energy transfer between modes. Note
        that $\nu = 0$ exactly changes the order of the PDE system, as the
        highest-order derivatives appear only in the dissipitive terms, so a
        nonzero $\nu$ must be chosen even in linear simulations.
\end{itemize}

@@

We aim to damp waves that reach the edge of the simulation domain without
inducing nonphysical reflection. To do so, we replace material derivatives
in \autoref{se:nl_upsilon} with:
\begin{align}
    \md{}{t} &\to \md{}{t} + \Gamma(z),\\
    \Gamma(z) &= \frac{1}{2\tau}\s{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}},\label{eq:Gamma}
\end{align}
where $z_B = 1.5H, z_T = 10.5H$ are the boundaries of the damping zone.
This strongly damps perturbations below $z_B$ and above $z_T$ with damping time
$\tau$, negligibly affects dynamics between $z_B, z_T$ and has transition width
governed by $\Delta z$. We use $\Delta z = 0.25H$ and $\tau = 1/15N$. This
prescription is similar to \citep{lecoanet_damp} and has the advantage of being
smooth, important for spectral methods. Further details of our implementation of
the fluid equations in Dedalus are described in Appendix~\ref{se:strat_impl}.

@@

We describe the results of a simulation with $C = \scinot{1.64}{-7}$ such that
$\xi_z k_z \ll 1$ everywhere. In particular, $\xi_z k_z \approx \scinot{5}{-5}$
just above the forcing zone. A nonzero $\nu$ corresponding to $\mathrm{Re} =
10^7$ was used.

We expect the waves to follow the solution given by \autoref{eq:uz_lin}. Along
with incompressibility (\autoref{eq:nl_incomp}), we obtain a complete analytical
solution for flow velocities $\bm{u}_{al}(x, z, t)$. The amplitude of the
observed IGW in the simulation field $\bm{u}$ relative to analytical solution
$\bm{u}_{al}$ over some region $z \in [z_b, z_t]$ can be estimated with
\begin{equation}
    A_i(t) = \frac{\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\p{\bm{u} \cdot \bm{u}_{al}}\;\mathrm{d}x\mathrm{d}z}
        {\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\abs{\bm{u}_{al}}^2\;\mathrm{d}x\mathrm{d}z}
        \label{eq:ahat_def}
\end{equation}
The subscript $i$ denotes the incident wave. If $\bm{u} = \bm{u}_{al}$, then
$A_i(t) = 1$. The normalization is chosen such that the overlap between $\bm{u},
\bm{u}_{al} \propto e^{z/2H}$ is evenly weighted throughout the integration
region.

For a simulation satisfying $\xi_z k_z \ll 1$ everywhere, we expect
$A_i(t) = 1$ when integrated between the forcing and damping zones, i.e.\
$z_b \gtrsim z_0, z_t \lesssim z_T$ ($z_0, z_T$ are defined in
\autoref{eq:vol_drive} and \autoref{eq:Gamma} respectively). For consistency
with the nonlinear case later, we choose $z_b = z_0 + 3\sigma, z_t = z_b + H$
(using $z_t = z_T - \Delta z$ just below the damping layer instead does not
change the results). The resulting measurement of $A_i(t)$ is shown in
\autoref{fig:lin_amps}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{plots/lin_amps.png}
    \caption{Amplitude of excited wave over time in weak forcing simulation,
    computed using \autoref{eq:ahat_def}. $A_i(t) = 1$ corresponds to
    perfect agreement with the analytical estimate. After an initial transient
    phase, we observe $A_i(t)$ asymptotes to $\approx 1$, implying continuous
    excitation of identical IGW with the expected amplitude. The small
    suppression may be due to timestepping errors, as a relatively large fixed
    step size $\Delta t = 0.1/N$ was used for this
    simulation.}\label{fig:lin_amps}
\end{figure}

@@

% times on fluxes.png are 054, 086, 153, 451

To perform simulations of wave breaking phenomena, we use the same values as
described in \autoref{s:numerics} except for $C$ and $\nu$. In particular, we
choose $C$ such that $\at{\xi_z k_z}_{z_0} = 0.1$ in the forcing zone. The
linear solution predicts $\xi_z k_z \sim 4.25$ at the upper damping zone $z_T$.
$\mathrm{Re}$ was chosen to be as large as possible across the various
resolutions. A table of our simulations can be found in \autoref{tab:params}.
\begin{table}
    \centering
    \begin{tabular}{l c c c}
        Resolution & $\mathrm{Re}$\\\bottomrule
        $1024 \times 4096$ & $2048$\\
        $768 \times 3072$ & $1024$\\
        $512 \times 2048$ & $512$\\
        $256 \times 1024$ & $341$\\
        $256 \times 1024$ & $205$\\
        $256 \times 1024$ & $146$\\
    \end{tabular}
    \caption{Table of simulation resolutions with wave
    breaking.}\label{tab:params}
\end{table}

@@

A full video of our simulation with $N_x = 768$, $N_z = 3072$, $\mathrm{Re} =
1024$ is available
online\footnote{http://www.princeton.edu/~lecoanet/data/breaking\_wave.mov}. We
take this to be our fiducial simulation for the remainder of this paper, though
other simulations show qualitatively similar behavior.

% convert yubo_000054.png -crop 2000x2000+150+250 out.png
In \autoref{fig:snapshots}, we present snapshots of $u_x$ and $\Upsilon$ at
various interesting phases of the simulation. The flow evolves through distinct
stages:
\begin{enumerate}
    \item At early times (top left panel), the flow resembles a linear IGW lower
        in the simulation domain but breaks down into smaller-scale features at
        higher $z$. Some characteristic swirling motion can be seen in the
        advected scalar $\Upsilon$, indicating Kelvin-Helmholtz instabilities.
